\documentclass[12pt,a4paper,twoside,openright,BCOR10mm,headsepline,titlepage,abstracton,chapterprefix,final]{scrreprt}

\usepackage{ae}
\usepackage[ngerman, english]{babel}
%\usepackage{SIunits}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{xcolor}
\usepackage{setspace}

% load hyperref as the last package to avoid redefinitions of e.g. footnotes after hyperref invokation

\RequirePackage{ifpdf}  % flag for pdf or dvi backend
\ifpdf
  \usepackage[pdftex]{graphicx}
  \usepackage[pdftitle={RTFM on Imaging Theory and Basics of Optical Raytracing},%
              pdfsubject={},%
              pdfauthor={M. Esslinger, J. Hartung, U. Lippmann},%
              pdfkeywords={},%
              bookmarks=true,%
%              colorlinks=true,%
              urlcolor=blue,%
              pdfpagelayout=TwoColumnRight,%
              pdfpagemode=UseNone,%
              pdfstartview=Fit,%
	      pdfpagelabels,
              pdftex]{hyperref}
\else
  \usepackage[dvips]{graphicx}
  \usepackage[colorlinks=false,dvips]{hyperref}
\fi
%\DeclareGraphicsRule{.jpg}{eps}{.jpg}{`convert #1 eps:-}

\usepackage{ae}
%\usepackage[ngerman, english]{babel}

%\usepackage{SIunits}
\newcommand\elementarycharge{\textrm{e}}
\newcommand\sccm{\textrm{sccm}}
\newcommand\mbar{\milli\textrm{bar}}


\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{setspace}

%\widowpenalty = 1000


\newcommand*{\doi}[1]{\href{https://doi.org/\detokenize{#1}}{doi: \detokenize{#1}}}

\newcommand\Vector[1]{{\mathbf{#1}}}

\newcommand\vacuum{0}

\newcommand\location{r}
\newcommand\Location{\Vector{r}}


\newcommand\wavenumber{k}
\newcommand\vacuumWavenumber{\wavenumber_{\vacuum}}
\newcommand\Wavevector{\Vector{\wavenumber}}

\newcommand\Nabla{\Vector{\nabla}}
\newcommand\Laplace{\Delta}
\newcommand\timederivative[1]{\dot{{#1}}}
\newcommand\Tensor[1]{\hat{#1}}
\newcommand\conjugate[1]{\overline{#1}}
\newcommand\transpose[1]{#1^{T}}
\newcommand\Norm[1]{\left| #1 \right|}
\newcommand{\ket}[1]{\left\vert{#1}\right\rangle}
\newcommand{\bra}[1]{\left\langle{#1}\right\vert}
\newcommand{\braket}[2]{\left\langle{#1}\vert{#2}\right\rangle}
\newcommand{\bracket}[1]{\left\langle{#1}\right\rangle}

\newcommand{\scpm}[2]{(#1\,\cdot\,#2)}

\newcommand\Greenfunction{\Tensor{G}}

\newcommand\scalarEfield{E}
\newcommand\scalarBfield{B}
\newcommand\scalarHfield{H}
\newcommand\scalarDfield{D}
\newcommand\scalarTipfield{T}
\newcommand\scalarSamplefield{S}
\newcommand\scalarDipolarmoment{p}
\newcommand\Efield{\Vector{\scalarEfield}}
\newcommand\Bfield{\Vector{\scalarBfield}}
\newcommand\Hfield{\Vector{\scalarHfield}}
\newcommand\Dfield{\Vector{\scalarDfield}}
\newcommand\Dipolarmoment{\Vector{\scalarDipolarmoment}}

\newcommand\permeability{\Tensor{\mu}}
\newcommand\vacuumpermeability{\mu_{\vacuum}}
\newcommand\permittivity{\Tensor{\epsilon}}
\newcommand\generalPermittivity{\Tensor{\tilde\epsilon}}
\newcommand\vacuumpermittivity{\epsilon_{\vacuum}}
\newcommand\scalarpermittivity{\epsilon}
\newcommand\conductivity{\Tensor{\sigma}}
\newcommand\susceptibility{\Tensor{\chi}}
\newcommand\currentdensity{\Vector{j}}
\newcommand\Current{\Vector{I}}
\newcommand\chargedensity{\rho}
\newcommand\PoyntingVector{\Vector{S}}

\newcommand\ordi{\text{ord}}
\newcommand\eo{\text{eo}}

\newcommand{\timeavg}[1]{{\langle\,#1\,\rangle}}

\newcommand{\remark}[1]{{\color{red}$\blacksquare$}\footnote{{\color{red}#1}}}
\newcommand{\chk}[1]{\color{green}{$\checkmark$#1}}


\newif\ifdraft
\draftfalse % \drafttrue




\begin{document}


\section{The Problem}
We are interested in solutions $x$ of the quartic equation
\begin{eqnarray}
 P = x^4 + a x^3 + b x^2 + c x + d = 0
\end{eqnarray}
where the coefficients $a,b,c,d$ are complex valued.

\section{Strategy}
Factor the equation into lower order terms.
There are several approaches how to do that.

\section{Direct Approach with 2 Parabolas}
The quartic equation shall be factored into two parabolic terms.
This approach does not depress the equation before solving.
It uses elementary math only.
\begin{eqnarray}
 x^4 + a x^3 + b x^2 + c x + d = ( x^2 + p x + q )( x^2 + rx + s) 
\end{eqnarray}
We compare coefficients
\begin{subequations}
  \begin{eqnarray}
    a &=& p + r
    \\
    b &=& s + q + pr
    \\
    c &=& ps + rq
    \\
    d &=& sq
  \end{eqnarray}
\end{subequations}
We need to solve this system of equations in order to factor the quartic equation. We consider the first 2 equations
\begin{subequations}
  \begin{eqnarray}
    r &=& a - p \\
    s &=& b - q - pr \\
      &=& b - q - p (a- p)
  \end{eqnarray}
\end{subequations}
and eliminate the coefficients of the second parabola $(r,s)$ in the latter 2 equations
\begin{subequations}
  \begin{eqnarray}
    c &=& p(b - q - p (a- p)) + (a-p)q
    \\
    d &=& (b - q - p (a- p))q
  \end{eqnarray}
\end{subequations}
We solve the $c$ equation for $q$
\begin{eqnarray}
 c &=& p(b - p (a- p)) + ( a- 2p )q \\
 q &=& \frac{c - p(b - p (a- p))}{a- 2p}
\end{eqnarray}
and insert in the equation for $d$
\begin{eqnarray}
    d &=& \left(b - \frac{c - p(b - p (a- p))}{a- 2p} - p (a- p)\right)\frac{c - p(b - p (a- p))}{a- 2p}
\end{eqnarray}
We reduced the system to one equation with $p$ as the only unknown.
\begin{eqnarray}
    d (a-2p)^2 &=& 
    \left(b(a-2p) - (c - p(b - p (a- p))) - p (a- p)(a-2p)\right)\cdot\nonumber\\
    &&\cdot\big(c - p(b - p (a- p))\big)
    \\
    &=& p^6 + ...
\end{eqnarray}
It is a polynomal of 6th degree in $p$.
This means there are 6 solutions for $p$.
We can understand the high degree considering the original polynomial as the product of its zeros $P=(x-x_1)(x-x_2)(x-x_3)(x-x_4)$.
Splitting up the polynomial in two parabolas leaves 6 possibilities.

\begin{tabular}{r|c|c}
       & $(x^2 + px + q)$ & $(x^2 + rx + s)$\\
 (i)   & $(x-x_1)(x-x_2)$ & $(x-x_3)(x-x_4)$\\
 (ii)  & $(x-x_1)(x-x_3)$ & $(x-x_2)(x-x_4)$\\
 (iii) & $(x-x_1)(x-x_4)$ & $(x-x_2)(x-x_3)$\\
 (iv)  & $(x-x_2)(x-x_3)$ & $(x-x_1)(x-x_4)$\\
 (v)   & $(x-x_2)(x-x_4)$ & $(x-x_1)(x-x_3)$\\
 (vi)  & $(x-x_3)(x-x_4)$ & $(x-x_1)(x-x_2)$\\
\end{tabular}

So, we transformed the problem of solving quartic polynomial into the seemingly more difficult problem of solving a polynomial of 6th degree. Fortunately, we can simplify the polynomial.
First we substitute $u=p(a-p)$, then we refactor the terms inside the first bracket on the right hand side.
\begin{eqnarray}
    d (a-2p)^2 &=& 
    \left(b(a-2p) - (c - p(b - u)) - u(a-2p)\right)
    \cdot
    \big(c - p(b - u)\big)
    \\
    &=& 
    \left((b-u)(a-2p) - c + p(b - u) \right)
    \cdot
    \big(c - p(b - u)\big)
    \\
    &=& 
    \left((b-u)(a-p) - c \right)
    \cdot
    \big(c - p(b - u)\big)
\end{eqnarray}
We further expand
\begin{eqnarray}
    d (a-2p)^2 &=& 
      (b-u)(a-p) c
      + p(b - u) c
      - c^2
      - (b-u)^2(a-p)p
    \\
    &=& (b-u)a - c^2 - u(b-u)^2
\end{eqnarray}
and also expand the left hand side
\begin{eqnarray}
    d (a^2 - 4ap + 4p^2) &=& (b-u)ac - c^2 - u(b-u)^2
    \\
    d (a^2 - 4u) &=& (b-u)ac - c^2 - u(b-u)^2
    \\
    d a^2 - 4 du &=& abc - acu - c^2 - u b^2 + 2bu^2 - u^3
\end{eqnarray}
and rearrange the terms
\begin{eqnarray}
   u^3 
   - (2b)u^2
   + ( ac + b^2 - 4 d ) u
   + (- abc + d a^2 + c^2)
   &=& 
   0
\end{eqnarray}
By solving this cubic equation, we can factor ans solve our quartic equation.


\subsection{Summary}
We can factor the quartic equation into two parabolas
\begin{eqnarray}
 x^4 + a x^3 + b x^2 + c x + d = ( x^2 + p x + q )( x^2 + rx + s) 
\end{eqnarray}
There are 6 solutions.
Any particular solution will factor the quartic equation.
\begin{eqnarray}
  p &=& \frac{a}{2} \pm \sqrt{ \frac{a^2}{4} - u}
\end{eqnarray}
where $u$ is a solution of the cubic equation
\begin{eqnarray}
  u^3 - (2b)u^2 + ( ac + b^2 - 4 d ) u + (- abc + d a^2 + c^2)
  &=& 0
\end{eqnarray}
The other coefficients follow the choice of the solution for $p$.
\begin{eqnarray}
 q &=& \frac{c - p(b - p (a- p))}{a- 2p} \\
 r &=& a - p \\
 s &=& \frac{d}{q}
\end{eqnarray}

\section{Application to Refraction}

\begin{align}
0 &=  \xi ^4 
      + \frac{c_3}{c_4} \xi ^3 
      + \frac{c_2}{c_4} \xi ^2 
      + \frac{c_1}{c_4} \xi 
      + \frac{c_0}{c_4}
\\
 &= ( \xi^2 + p \xi + q )( \xi^2 + r \xi + s) 
\end{align}
\begin{eqnarray}
  p &=& \frac{c_3}{2 c_4} \pm \sqrt{ \frac{c_3^2}{4 c_4^2} - x}
\end{eqnarray}
where $x$ is a solution of the cubic equation
\begin{eqnarray}
  x^3 
  - \frac{2 c_2}{c_4} x^2 
  + \frac{c_1 c_3 + c_2^2 - 4 c_0 c_4}{c_4^2} x
  + \frac{- c_3 c_2 c_1 + c_0 c_3^2 + c_1^2 c_4}{c_4^3}
  &=& 0
\end{eqnarray}
we abbreviate the polynomial coefficients
\begin{eqnarray}
 0 &=& x^3 + b x^2 + c x + d\\
 b &=& - \frac{2 c_2}{c_4}\\
 c &=&   \frac{c_1 c_3 + c_2^2 - 4 c_0 c_4}{c_4^2}\\
 d &=&   \frac{- c_3 c_2 c_1 + c_0 c_3^2 + c_1^2 c_4}{c_4^3}
\end{eqnarray}
We perform a linear Tschirnhaus transformation
$x = z -\frac{b}{3}$ to depress the equation
\begin{eqnarray}
0 &=& z^3 + \tilde{p} z + \tilde{q}
\\
\tilde{p} &=& 3 e^2 + 2 \frac{b}{a} e + \frac{c}{a}
\\
\tilde{q} &=& e^3 + \frac{b}{a} e^2 + \frac{c}{a} e + \frac{d}{a}
\\
e &=& - \frac{b}{3}
\end{eqnarray}



\section{Application to Surface Intersections}
Such equations are in principle analytically solvable. But on the one hand it is very complicated and on the other hand 
it is not known whether there exists such a user-friendly numerically stable form of $t$ like in the former cases.\footnote{
In fact the word ``biconic" is misleading because it suggests that
it is a surface of second degree which is wrong. \eqref{eq:implicitbiconic}
shows clearly that it is a surface of degree four.}

Let us first assume that $\Vector{r} = \Vector{r}_0 + t \Vector{d}$ is the insertion of the ray into
\eqref{eq:implicitbiconic}. Then $a(x(t), y(t)) =: Q_2(t)$ and $b(x(t), y(t)) =: Q_1(t)$ become quadratic polynomials 
in $t$. Further $z(t) =: L(t)$ becomes linear in $t$. Therefore this leads to
\begin{align}
 (Q_2(t) - 1) \underbrace{L^2(t)}_{\ge0} + \underbrace{(L(t) - Q_1(t))^2}_{\ge0} &= 0\,.\label{eq:biconict}
\end{align}
From \eqref{eq:biconic} and \eqref{eq:implicitbiconic} it follows that $1 - Q_2(t) \ge 0$ such that the
prefactor in \eqref{eq:biconict} is non-positive. Therefore the intersection equations lead to a real
fourth order polynomial which has to be solved. The quadratic polynomial $Q_2(t) - 1$ can be factorized
\begin{align}
 Q_2(t) - 1 &= A_2 (t - t_{Q21})(t - t_{Q22}) \le 0\,,
\end{align}
where $A_2 = (1 + c_x) \rho_x^2 d_x^2 + (1 + c_y) \rho_y^2 d_y^2$. If $A_2 > 0$ then at least one zero must exist and
$t \in [t_{Q21}, t_{Q22}]$. If $A_2 = 0$ then $Q_2(t)$ is not quadratic anymore. If $A_2 < 0$ then in 
\begin{align}
 Q_2(t) - 1 &= A_2 \left(\left(t-\frac{t_{Q21} + t_{Q22}}{2}\right)^2 - \frac{1}{4}(t_{Q21} - t_{Q22})^2\right)\,,
\end{align}
the expression in brackets has to be positive or zero. This is only true if the zeros are either the same or they are complex, because
for $t_{Q21/2} = \tau \pm i \sigma$ this expression becomes
\begin{align}
 Q_2(t) - 1 &= A_2 \left(\left(t-\tau\right)^2 + \frac{1}{2}\sigma^2\right)<0\,,
\end{align}
for real $t$. However the final equation is given by
\begin{align}
 a t^4 + b t^3 + c t^2 + d t + e &= 0\,,
\end{align}
where\remark{Simplify coefficients!}
\begin{subequations}
 \begin{align}
  a &= c_{x}^{2} d_{x}^{4} + 2 \, c_{x} c_{y} d_{x}^{2} d_{y}^{2} + c_{y}^{2} d_{y}^{4} \nonumber\\&
       + {\left(c_{x}^{2} k_{x} + c_{x}^{2}\right)} d_{x}^{2} d_{z}^{2} + {\left(c_{y}^{2} k_{y} + c_{y}^{2}\right)} d_{y}^{2} d_{z}^{2}\,,\\
  b &= 4 \, c_{x}^{2} d_{x}^{3} x_{0} + 4 \, c_{x} c_{y} d_{x} d_{y}^{2} x_{0} + 4 \, c_{x} c_{y} d_{x}^{2} d_{y} y_{0} + 4 \, c_{y}^{2} d_{y}^{3} y_{0}\nonumber\\&
       + 2 \, {\left(c_{x}^{2} k_{x} + c_{x}^{2}\right)} d_{x} d_{z}^{2} x_{0} + 2 \, {\left(c_{y}^{2} k_{y} + c_{y}^{2}\right)} d_{y} d_{z}^{2} y_{0}\nonumber\\&
       + 2 \, {\left(c_{x}^{2} k_{x} + c_{x}^{2}\right)} d_{x}^{2} d_{z} z_{0} + 2 \, {\left(c_{y}^{2} k_{y} + c_{y}^{2}\right)} d_{y}^{2} d_{z} z_{0}\nonumber\\&
       - 2 \, c_{x} d_{x}^{2} d_{z} - 2 \, c_{y} d_{y}^{2} d_{z}\,,\\
  c &= 6 \, c_{x}^{2} d_{x}^{2} x_{0}^{2} + 2 \, c_{x} c_{y} d_{y}^{2} x_{0}^{2} + 8 \, c_{x} c_{y} d_{x} d_{y} x_{0} y_{0} + 2 \, c_{x} c_{y} d_{x}^{2} y_{0}^{2}\nonumber\\&
     + 6 \, c_{y}^{2} d_{y}^{2} y_{0}^{2} + {\left(c_{x}^{2} k_{x} + c_{x}^{2}\right)} d_{z}^{2} x_{0}^{2} + {\left(c_{y}^{2} k_{y} + c_{y}^{2}\right)} d_{z}^{2} y_{0}^{2}\nonumber\\&
     + 4 \, {\left(c_{x}^{2} k_{x} + c_{x}^{2}\right)} d_{x} d_{z} x_{0} z_{0} + 4 \, {\left(c_{y}^{2} k_{y} + c_{y}^{2}\right)} d_{y} d_{z} y_{0} z_{0}\nonumber\\&
     + {\left(c_{x}^{2} k_{x} + c_{x}^{2}\right)} d_{x}^{2} z_{0}^{2} + {\left(c_{y}^{2} k_{y} + c_{y}^{2}\right)} d_{y}^{2} z_{0}^{2}\nonumber\\&
     - 4 \, c_{x} d_{x} d_{z} x_{0} - 4 \, c_{y} d_{y} d_{z} y_{0} - 2 \, c_{x} d_{x}^{2} z_{0} - 2 \, c_{y} d_{y}^{2} z_{0}\,,\\
  d &= 4 \, c_{x}^{2} d_{x} x_{0}^{3} + 4 \, c_{x} c_{y} d_{y} x_{0}^{2} y_{0}\nonumber\\&
     + 4 \, c_{x} c_{y} d_{x} x_{0} y_{0}^{2} + 4 \, c_{y}^{2} d_{y} y_{0}^{3}\nonumber\\&
     + 2 \, {\left(c_{x}^{2} k_{x} + c_{x}^{2}\right)} d_{z} x_{0}^{2} z_{0} + 2 \, {\left(c_{y}^{2} k_{y} + c_{y}^{2}\right)} d_{z} y_{0}^{2} z_{0} 
     + 2 \, {\left(c_{x}^{2} k_{x} + c_{x}^{2}\right)} d_{x} x_{0} z_{0}^{2} + 2 \, {\left(c_{y}^{2} k_{y} + c_{y}^{2}\right)} d_{y} y_{0} z_{0}^{2}\nonumber\\&
     - 2 \, c_{x} d_{z} x_{0}^{2} - 2 \, c_{y} d_{z} y_{0}^{2} - 4 \, c_{x} d_{x} x_{0} z_{0} - 4 \, c_{y} d_{y} y_{0} z_{0}\,,\\
  e &= c_{x}^{2} x_{0}^{4} + 2 \, c_{x} c_{y} x_{0}^{2} y_{0}^{2} + c_{y}^{2} y_{0}^{4} + {\left(c_{x}^{2} k_{x} + c_{x}^{2}\right)} x_{0}^{2} z_{0}^{2}\nonumber\\&
    + {\left(c_{y}^{2} k_{y} + c_{y}^{2}\right)} y_{0}^{2} z_{0}^{2} - 2 \, c_{x} x_{0}^{2} z_{0} - 2 \, c_{y} y_{0}^{2} z_{0}\,.
 \end{align}
\end{subequations}
Then we may consider the nature of the zeros (see Wiki) by considering the following combinations, namely
the discriminant $\Delta$, the quadratic coefficient of the depressed quartic $P$, the linear coefficient of
the depressed quartic $Q$, $\Delta_0$ which is zero if the quartic has a triple root and $D$ which is zero
if the quartic has two double roots:
\begin{subequations}
 \begin{align}
  \Delta &= 256 a^3 e^3 - 192 a^2 b d e^2 - 128 a^2 c^2 e^2 + 144 a^2 c d^2 e - 27 a^2 d^4 \nonumber\\ 
&+ 144 a b^2 c e^2 - 6 a b^2 d^2 e - 80 a b c^2 d e + 18 a b c d^3 + 16 a c^4 e \nonumber\\
&- 4 a c^3 d^2 - 27 b^4 e^2 + 18 b^3 c d e - 4 b^3 d^3 - 4 b^2 c^3 e + b^2 c^2 d^2\,,\\
   P &= 8ac - 3b^2\,,\\
   Q &= b^3+8da^2-4abc\,,\\
   \Delta_0 &= c^2 - 3bd + 12ae\,,\\
   D &= 64 a^3 e - 16 a^2 c^2 + 16 a b^2 c - 16 a^2 bd - 3 b^4\,.
 \end{align}
\end{subequations}

\begin{itemize}
\item If $\Delta < 0$  then the equation has two real roots and two complex conjugate roots.
\item If $\Delta > 0$  then the equation's four roots are either all real or all complex.
\begin{itemize}
\item If $P < 0$ and $D < 0$ then all four roots are real and distinct.
\item If $P > 0$ or $D > 0$ then there are two pairs of complex conjugate roots.
\end{itemize}
\item If $\Delta = 0$  then either the polynomial has a multiple root, 
  or it is the square of a quadratic polynomial. Here are the different cases that can occur:
  \begin{itemize}
\item If $P < 0$ and $D < 0$ and $\Delta_0\ne0$, there is a real double root and two real simple roots.
\item If $D > 0$ or ($P > 0$ and ($D \ne 0$ or $Q \ne 0$)), there is a real double root and two complex conjugate roots.
\item If $\Delta_0 = 0$ and $D \ne 0$, there is a triple root and a simple root, all real.
\item If $D = 0$, then:
  \begin{itemize}
\item If $P < 0$, there are two real double roots.
\item If $P > 0$ and $Q = 0$, there are two complex conjugate double roots.
\item If $ \Delta_0  = 0$, all four roots are equal to $-\frac{b}{4a}$
  \end{itemize}
  \end{itemize}
\end{itemize}
Some of the discussed cases show that the ray misses the surfaces.



\end{document}
