\documentclass[12pt,a4paper,twoside,openright,BCOR10mm,headsepline,titlepage,abstracton,chapterprefix,final]{scrreprt}

\usepackage{ae}
\usepackage[ngerman, english]{babel}
%\usepackage{SIunits}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{xcolor}
\usepackage{setspace}
\usepackage{dsfont}

% load hyperref as the last package to avoid redefinitions of e.g. footnotes after hyperref invokation

\RequirePackage{ifpdf}  % flag for pdf or dvi backend
\ifpdf
  \usepackage[pdftex]{graphicx}
  \usepackage[pdftitle={RTFM on Imaging Theory and Basics of Optical Raytracing},%
              pdfsubject={},%
              pdfauthor={M. Esslinger, J. Hartung, U. Lippmann},%
              pdfkeywords={},%
              bookmarks=true,%
%              colorlinks=true,%
              urlcolor=blue,%
              pdfpagelayout=TwoColumnRight,%
              pdfpagemode=UseNone,%
              pdfstartview=Fit,%
	      pdfpagelabels,
              pdftex]{hyperref}
\else
  \usepackage[dvips]{graphicx}
  \usepackage[colorlinks=false,dvips]{hyperref}
\fi
%\DeclareGraphicsRule{.jpg}{eps}{.jpg}{`convert #1 eps:-}

\usepackage{ae}
%\usepackage[ngerman, english]{babel}

%\usepackage{SIunits}
\newcommand\elementarycharge{\textrm{e}}
\newcommand\sccm{\textrm{sccm}}
\newcommand\mbar{\milli\textrm{bar}}


\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{setspace}

%\widowpenalty = 1000


\newcommand*{\doi}[1]{\href{http://dx.doi.org/\detokenize{#1}}{doi: \detokenize{#1}}}

\newcommand\Vector[1]{{\mathbf{#1}}}

\newcommand\vacuum{0}

\newcommand\location{r}
\newcommand\Location{\Vector{r}}


\newcommand\wavenumber{k}
\newcommand\vacuumWavenumber{\wavenumber_{\vacuum}}
\newcommand\Wavevector{\Vector{\wavenumber}}

\newcommand\Nabla{\Vector{\nabla}}
\newcommand\Laplace{\Delta}
\newcommand\timederivative[1]{\dot{{#1}}}
\newcommand\Tensor[1]{\hat{#1}}
\newcommand\conjugate[1]{\overline{#1}}
\newcommand\transpose[1]{#1^{T}}
\newcommand\Norm[1]{\left| #1 \right|}
\newcommand{\ket}[1]{\left\vert{#1}\right\rangle}
\newcommand{\bra}[1]{\left\langle{#1}\right\vert}
\newcommand{\braket}[2]{\left\langle{#1}\vert{#2}\right\rangle}
\newcommand{\bracket}[1]{\left\langle{#1}\right\rangle}

\newcommand{\orderof}[1]{\mathcal{O}(#1)}

\newcommand{\scpm}[2]{(#1\,\cdot\,#2)}

\newcommand\Greenfunction{\Tensor{G}}

\newcommand\scalarEfield{E}
\newcommand\scalarBfield{B}
\newcommand\scalarHfield{H}
\newcommand\scalarDfield{D}
\newcommand\scalarTipfield{T}
\newcommand\scalarSamplefield{S}
\newcommand\scalarDipolarmoment{p}
\newcommand\Efield{\Vector{\scalarEfield}}
\newcommand\Bfield{\Vector{\scalarBfield}}
\newcommand\Hfield{\Vector{\scalarHfield}}
\newcommand\Dfield{\Vector{\scalarDfield}}
\newcommand\Dipolarmoment{\Vector{\scalarDipolarmoment}}

\newcommand\permeability{\Tensor{\mu}}
\newcommand\vacuumpermeability{\mu_{\vacuum}}
\newcommand\permittivity{\Tensor{\epsilon}}
\newcommand\generalPermittivity{\Tensor{\tilde\epsilon}}
\newcommand\vacuumpermittivity{\epsilon_{\vacuum}}
\newcommand\scalarpermittivity{\epsilon}
\newcommand\conductivity{\Tensor{\sigma}}
\newcommand\susceptibility{\Tensor{\chi}}
\newcommand\currentdensity{\Vector{j}}
\newcommand\Current{\Vector{I}}
\newcommand\chargedensity{\rho}
\newcommand\PoyntingVector{\Vector{S}}

\newcommand{\remark}[1]{{\color{red}$\blacksquare$}\footnote{{\color{red}#1}}}

\newif\ifdraft
\draftfalse % \drafttrue




\begin{document}

\section{Pilot Ray Concept}

\subsection{Motivation}

In systems like Offner systems or off-axis parabolic mirrors, the center-of-field chief ray is no more aligned with the axis of rotational symmetry of the surfaces.
In these systems, it is no longer sufficient to obtain first order properties from the central curvature on the axis of symmetry of each surface.
In other systems, like freeshape Schiefspiegler telescopes, there is no axis of symmetry at all. 
Anamorphotic systems, on the other hand, often have an optical axis, but do not have the same effective surface curvature in all meridional planes.

In this section, we develop a description for the first order properties of optical systems 
and describe rotationally symmetric and non-symmetric systems with a single formalism.
It is a formalism for rays close to the center-of-field chief ray, which we call \emph{pilot ray} in the following.
This pilot ray takes the role of the optical axis in rotationally symmetric systems. 
By choosing the term pilot ray, we want to emphasize that we mean neither a mechanical axis of symmetry of individual lens elements nor the extraordinary axis in anisotropic crystals.
In most systems, the intial pilot ray direction is perpendicular to the object plane. 
An exception is found, for example, in Scheimpflug systems.

Our ansatz is to perform a real ray trace of the pilot ray and describe a small area around it by a Taylor expansion around the real ray intersection point.
From this Taylor expansion, we derive first order properties, like focal length, magnification and image position.
The concept is a parabasal approximation around a non-paraxial pilot ray.



\subsection{Near-Pilot Rays}

We consider near-pilot rays, that have an intersection point and a wavevector close to the pilot ray.

\begin{tabular}{ l | l | l }
		    & pilot ray & near-pilot ray \\
& & \\ \hline & & \\
intersection point  & $\Vector{r}_{pilot}$ & $\tilde{\Vector{r}} = \Vector{r}_{pilot} + \Delta\Vector{r}$ \\
initial wavevector  & $\Wavevector_{pilot}$ & $\tilde{\Wavevector} = \Wavevector_{pilot} + \Delta\Wavevector$ \\
initial ray direction  & $\Vector{d}_{pilot}$ & $\tilde{\Vector{d}} = \Vector{d}_{pilot} + \Delta\Vector{d}$ \\
surface normal      & $\Vector{n}_{pilot} = \Vector{n}(\Vector{r}_{pilot})$ & $\tilde{\Vector{n}} = \Vector{n}(\tilde{\Vector{r}}) = \Vector{n}_{pilot} + \Delta\Vector{n}$ \\
& & \\ \hline & & \\
out-of-plane wavevector component &  $\Wavevector_{pilot, \perp} = (\Wavevector_{pilot} \Vector{n})\Vector{n}$ & $\tilde{\Wavevector}_{\perp} = (\tilde{\Wavevector} \tilde{\Vector{n}})\tilde{\Vector{n}}$ \\
in-plane wavevector component     &  $\Wavevector_{pilot, \parallel} = \Wavevector_{pilot} - \Wavevector_{pilot, \perp}$ & $\tilde{\Wavevector}_{\parallel} = \tilde{\Wavevector} - \tilde{\Wavevector}_{\perp}$ \\
\end{tabular}
\\[2ex]

\subsubsection{Representation of near-pilot ray directions}

There is an additional constraint on the ray direction unit vectors: both pilot and near-pilot ray direction must be normalized,
$|\Vector{d}_{pilot}|=|\tilde{\Vector{d}}|=1$.

\begin{eqnarray}
 |\tilde{\Vector{d}}|^2 = |\Vector{d}_{pilot}|^2 + 2 \Vector{d}_{pilot} \cdot \Delta\Vector{d} + |\Delta\Vector{d}|^2
\end{eqnarray}

In the limit of near pilot rays, $|\Delta\Vector{d}|^2 \rightarrow 0$, the scalar product must vanish. 
The change of direction is perpendicular to the pilot direction, $\Vector{d}_{pilot} \perp \Delta\Vector{d}$. 
This leaves only two degrees of freedom for $\Delta\Vector{d}$.

\begin{eqnarray}
 \Delta\Vector{d} &=& U \Vector{e}_{U} + V \Vector{e}_{V}
\end{eqnarray}

We represent $\Delta\Vector{d}$ by two components $U$and $V$ and choose orthogonal unit vectors $\Vector{e}_{U} \perp \Vector{e}_{V} \perp \Vector{d}_{pilot} \perp \Vector{e}_{U}$,
forming a cartesian coordinate system.
In this representation, the normalization requirement $|\tilde{\Vector{d}}| = 1$ imposes no nonlinear constraint on a third component depending on the choice of $U$ and $V$.
We may choose (small) values for $U$ and $V$ independent from each other and independent from the pilot direction. 

\subsubsection{Representation of near-pilot intersection points}

In parabasal approximation, we assume each surface planar in a small area around the pilot ray intersection point. 
We represent the ray height in this plane by

\begin{eqnarray}
 \Delta\Vector{r} = X \Vector{e}_{X} + Y \Vector{e}_{Y}
\end{eqnarray}

where $\Vector{e}_{X}$ and $\Vector{e}_{Y}$ are in the tangential surface plane.
As we will see later, this choice is beneficial when calculating near-pilot ray intersection points.
The disadvantage of this representation is that it is not identical to the coordinate system used to describe the surface sag.
In general, the planes for $\Delta\Vector{d}$ and $\Delta\Vector{r}$ may differ.
For simplicity, we choose the two systems $XY$ and $UV$ identical,
except for a component along the pilot ray axis.

\begin{eqnarray}
 \Vector{e}_{U} &=& \frac{\Vector{e}_{X} - (\Vector{e}_{X}\Vector{d}_{pilot})\Vector{d}_{pilot}}{|\Vector{e}_{X} - (\Vector{e}_{X}\Vector{d}_{pilot})\Vector{d}_{pilot}|}
 \\[2ex]
 \Vector{e}_{V} &=& \frac{\Vector{e}_{Y} - (\Vector{e}_{Y}\Vector{d}_{pilot})\Vector{d}_{pilot}}{|\Vector{e}_{Y} - (\Vector{e}_{Y}\Vector{d}_{pilot})\Vector{d}_{pilot}|}
\end{eqnarray}

With this choice, we restrict ourselves to cases where the pilot direction does not lie inside the plane of $\Vector{e}_{X}$ and $\Vector{e}_{Y}$.

\subsection{Unique choice of unit vectors}
Up to now, the coordinate systems $\Vector{e}_{X}, \Vector{e}_{Y}$ are defined except for a \emph{roll} degree of freedom, that is, a rotation around the pilot ray.
In case we partition our optical system into subsets, for example individual optical elements or surfaces, each subsystem $i$ may have its own coordinate system
$\Vector{e}_{X,i}, \Vector{e}_{Y,i}$.
Each subsystem may have an arbitrarily chosen roll angle. 
There is, among others, one choices we want to highlight:

We take the local coordinate system used to describe the surface sag and project its unit vectors on the tangential plane.
The projections of sag $x$ and $y$ unit vectors are then employed for $X$ and $Y$ ray heights, respectively.
With this choice, we can easily express absolute vertex heights of near pilot rays back in the sag coordinate system. 
In case the pilot ray lies in the sag $xy$ plane, however, we face numerical difficulties.

\begin{eqnarray}
 \Vector{e}_{X} &=& \frac{\Vector{e}_{x} - (\Vector{e}_{x}\Vector{n}_{pilot})\Vector{n}_{pilot}}{|\Vector{e}_{x} - (\Vector{e}_{x}\Vector{n}_{pilot})\Vector{n}_{pilot}|}
 \\[2ex]
 \Vector{e}_{Y} &=& \frac{\Vector{e}_{y} - (\Vector{e}_{y}\Vector{n}_{pilot})\Vector{n}_{pilot}}{|\Vector{e}_{y} - (\Vector{e}_{y}\Vector{n}_{pilot})\Vector{n}_{pilot}|}
\end{eqnarray}


\subsection{Propagation of rays through an optical system}

After tracing the pilot ray, we also want to propagate near-pilot rays.

\begin{eqnarray}
 \begin{pmatrix}
  X_{im} \\ Y_{im} \\ U_{im} \\ V_{im}
 \end{pmatrix}
 &=&
 \hat{G}
 \begin{pmatrix}
  X_{obj} \\ Y_{obj} \\ U_{obj} \\ V_{obj}
 \end{pmatrix}
\end{eqnarray}

The coordinates $X,Y,U,V$ reside in their respective object and image sided coordinate systems $\Vector{e}$.
By definition, the propagator $\hat{G}$ projects the object sided pilot ray onto the image sided pilot ray.
$\hat{G}\Vector{0} = \Vector{0}$.

In analogy to the ABCD formalism, 
\begin{eqnarray}
 \begin{pmatrix}
  y_{im} \\ y^{\,\prime}_{im}
 \end{pmatrix}
 =
 \begin{pmatrix}
   A & B \\ C & D
 \end{pmatrix}
 \begin{pmatrix}
   y_{obj} \\ y^{\,\prime}_{obj}
 \end{pmatrix}
\end{eqnarray}
we neglect terms quadratic in the change of direction or position, 
$\orderof{\Delta\Vector{d}^2}$ and $\orderof{\Delta\Vector{r}^2}$, as well as mixed terms $\orderof{\Delta\Vector{d}\Delta\Vector{r}}$.
The linear terms in $\hat{G}$ can be represented by a matrix.

\begin{eqnarray}
 \hat{G} &=&
 \begin{pmatrix}
  a_{11} & a_{12} & b_{11} & b_{12} \\
  a_{21} & a_{22} & b_{21} & b_{22} \\  
  c_{11} & c_{12} & d_{11} & d_{12} \\
  c_{21} & c_{22} & d_{21} & d_{22} 
 \end{pmatrix}
 =
 \begin{pmatrix}
  \hat{A} & \hat{B} \\
  \hat{C} & \hat{D}
 \end{pmatrix} 
\end{eqnarray}

Like in the ABCD-formalism, we can express the effect of multiple propagators as the product of its matrices.
The individual propagators can be, for example, free space propagation or surface refraction.

\begin{eqnarray}
 \begin{pmatrix}
  X_{im} \\ Y_{im} \\ U_{im} \\ V_{im}
 \end{pmatrix}
 &=&
 \hat{G}_N \cdot \hat{G}_{N-1} \cdot ... \cdot \hat{G}_3 \cdot \hat{G}_2 \cdot \hat{G}_1 \cdot
 \begin{pmatrix}
  X_{obj} \\ Y_{obj} \\ U_{obj} \\ V_{obj}
 \end{pmatrix}
\end{eqnarray}

In general, the matrices do not commutate. $\hat{G}_1$ is the propagator closest to the object.


\subsection{Free Space Propagation}
In this section, we discuss propagation of straight rays on the tangential plane of the following surface.
Rays travel along straights in homogeneous media.

Assume we have a pilot ray with starting point $\Vector{r}_{pilot,0}$ and a near-pilot ray $\tilde{\Vector{r}_0}$.
First, we propagate the pilot ray to its intersection point $\Vector{r}_{pilot,1}$ and then the near pilot ray.
\begin{eqnarray}
 %\Vector{d}_{pilot,1} &=& \Vector{d}_{pilot,0} \\
 %\Vector{r}_{pilot,1} &=& \Vector{r}_{pilot,0} + \Vector{d}_{pilot,0} t_{pilot} \\
 \tilde{\Vector{d}}_1 &=& \tilde{\Vector{d}}_0 \\
 \tilde{\Vector{r}}_1 &=& \tilde{\Vector{r}}_0 + \tilde{\Vector{d}}_0 t \\
 \tilde{\Vector{r}}_1 &=& \Vector{r}_{pilot,1} + X_1 \Vector{e}_{X1} + Y_1 \Vector{e}_{Y1}
\end{eqnarray}
The free parameter of the straight $t$ has to be chosen to intersect surface $1$.
\begin{eqnarray}
 \tilde{\Vector{r}}_{pilot,0} - \Vector{r}_{pilot,1} + X_0 \Vector{e}_{X0} + Y_0 \Vector{e}_{Y0} 
 &=& 
 X_1 \Vector{e}_{X1} + Y_1 \Vector{e}_{Y1} - \tilde{\Vector{d}}_0 t
\end{eqnarray}
We have a 3 dimensional vectorial equation and 3 unknowns $X_1, Y_1, t$.
In cartesian coordinates, it reads
\begin{eqnarray}
 \begin{pmatrix}
  X_1 \\ Y_1 \\ t
 \end{pmatrix}
 &=&
 \begin{pmatrix}
  e_{X1x} & e_{Y1x} & d_{0x} \\
  e_{X1y} & e_{Y1y} & d_{0y} \\
  e_{X1z} & e_{Y1z} & d_{0z}
 \end{pmatrix}^{-1}
 \left(
  \tilde{\Vector{r}}_{pilot,0} - \Vector{r}_{pilot,1} + X_0 \Vector{e}_{X0} + Y_0 \Vector{e}_{Y0} 
 \right)
\end{eqnarray}

\subsection{Thin lenses}
We consider a thin, spherical lens in air with focal length $f$ and no tilt of the pilot ray, 
$\Vector{e}_{X1} = \Vector{e}_{X0}$ and $\Vector{e}_{Y1} = \Vector{e}_{Y0}$. Its XYUV matrix reads

\begin{eqnarray}
XYUV_{thin lens} &=&
 \begin{pmatrix}
  1 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 \\
  -\frac{1}{f} & 0 & 1 & 0 \\
  0 & -\frac{1}{f} & 0 & 1
 \end{pmatrix}
\end{eqnarray}

and for a cylindrical lens with a focal power tilted by $\theta$ with respect to the $Y$ axis

\begin{eqnarray}
XYUV_{cyl} &=&
 \begin{pmatrix}
  1 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 \\
  -\frac{1}{f} \sin^2\theta & -\frac{1}{f} \sin\theta \cos\theta & 1 & 0 \\
  -\frac{1}{f} \sin\theta \cos\theta & -\frac{1}{f} \cos^2\theta & 0 & 1
 \end{pmatrix}
\end{eqnarray}



\subsection{Evaluation of the Law of Refraction}



TODO: adapt to new nomenclature

We are interested in the change of the refracted wavevector $\Delta\Wavevector_2 = \tilde{\Wavevector}_2 - \Wavevector_2$ induced by $\Delta\Vector{r}$ and $\Delta\Wavevector$.
\begin{eqnarray}
 \Wavevector_2 &=& \Wavevector_{\parallel} + \xi \Vector{n} 
 \\
 \tilde{\Wavevector}_2 &=& \tilde{\Wavevector}_{\parallel} + \tilde{\xi} \tilde{\Vector{n}} 
\end{eqnarray}

The refraction conserves the in-plane wavevector component. 
The out-of-plane or normal wavevector component $\xi, \tilde{\xi}$ after the refraction depends on the material permittivity and the in-plane component.

\begin{eqnarray}
 \Delta\Wavevector_2 &=& \tilde{\Wavevector}_2 - \Wavevector_2 \\
 &=& ( \tilde{\Wavevector}_{\parallel} - \Wavevector_{\parallel} ) + ( \tilde{\xi} \tilde{\Vector{n}} - \xi \Vector{n} ) \\
% &=& ( \tilde{\Wavevector} - \Wavevector)  - (\tilde{\Wavevector}_{\perp} - \Wavevector_{\perp}) + ( \tilde{\xi} - \xi ) \Vector{n} + \tilde{\xi} \Delta\Vector{n} \\
% &=& \Delta\Wavevector - (\tilde{\Wavevector}_{\perp} - \Wavevector_{\perp}) + ( \tilde{\xi} - \xi ) \Vector{n} + \tilde{\xi} \Delta\Vector{n} \\
% &=& \Delta\Wavevector - ((\tilde{\Wavevector} \tilde{\Vector{n}})\tilde{\Vector{n}} - (\Wavevector \Vector{n})\Vector{n}) + ( \tilde{\xi} - \xi ) \Vector{n} + \tilde{\xi} \Delta\Vector{n} \\
% &=& \Delta\Wavevector - 
% \left(
%   (\tilde{\Wavevector} \tilde{\Vector{n}})\Vector{n}+(\tilde{\Wavevector} \tilde{\Vector{n}})\Delta\Vector{n} - (\Wavevector \Vector{n})\Vector{n}
% \right) 
% + ( \tilde{\xi} - \xi ) \Vector{n} + \tilde{\xi} \Delta\Vector{n} \\
% &=& \Delta\Wavevector - 
% \left(
%   (\tilde{\Wavevector} (\Vector{n} + \Delta\Vector{n}) )\Vector{n}+(\tilde{\Wavevector} (\Vector{n} + \Delta\Vector{n}) )\Delta\Vector{n} - (\Wavevector \Vector{n})\Vector{n}
% \right) 
% + ( \tilde{\xi} - \xi ) \Vector{n} + \tilde{\xi} \Delta\Vector{n} \\
% &=& \Delta\Wavevector - 
% \left(
%   (\tilde{\Wavevector} \Vector{n} )\Vector{n} + (\tilde{\Wavevector} \Delta\Vector{n} )\Vector{n} + (\tilde{\Wavevector} \Vector{n} )\Delta\Vector{n} + (\tilde{\Wavevector} \Delta\Vector{n} )\Delta\Vector{n} - (\Wavevector \Vector{n})\Vector{n}
% \right) 
% + ( \tilde{\xi} - \xi ) \Vector{n} + \tilde{\xi} \Delta\Vector{n} \\
% &=& \Delta\Wavevector - 
% \left(
%   (\Delta\Wavevector \Vector{n} )\Vector{n} + (\tilde{\Wavevector} \Delta\Vector{n} )\Vector{n} + (\tilde{\Wavevector} \Vector{n} )\Delta\Vector{n} + (\tilde{\Wavevector} \Delta\Vector{n} )\Delta\Vector{n}
% \right) 
% + ( \tilde{\xi} - \xi ) \Vector{n} + \tilde{\xi} \Delta\Vector{n} \\
 &=& \Delta\Wavevector - 
 \bigg(
   (\Delta\Wavevector \Vector{n} )\Vector{n} 
   + (\Wavevector \Delta\Vector{n} )\Vector{n}  + (\Delta\Wavevector \Delta\Vector{n} )\Vector{n} 
   + (\Wavevector \Vector{n} )\Delta\Vector{n}  + 
  \nonumber \\ &&
   + (\Delta\Wavevector \Vector{n} )\Delta\Vector{n} 
   + (\Wavevector \Delta\Vector{n} )\Delta\Vector{n}
   + (\Delta\Wavevector \Delta\Vector{n} )\Delta\Vector{n}
 \bigg) 
 + ( \tilde{\xi} - \xi ) \Vector{n} + \tilde{\xi} \Delta\Vector{n}
\end{eqnarray}
We are insterested in the limit $\Delta\Wavevector, \Delta\Vector{r} \rightarrow 0$. We approximate $\Delta\Vector{n}$ in first order
\begin{eqnarray}
 \Delta n_i \approx \left. \frac{\partial n_i}{\partial r_j} \right|_{\Vector{r}_{pilot}} \Delta r_j
\end{eqnarray}
to be proportional to $\Delta\Vector{r}$. 
We neglect all higher order terms $\Delta\Wavevector^2$ and $\Delta\Vector{r}^2$.
We also neglect cross-terms $\Delta\Wavevector\Delta\Vector{r}$. 
This leaves the lowest order terms
\begin{eqnarray}
% \Delta\Wavevector_2 &\approx& 
%   \Delta\Wavevector 
%   - (\Delta\Wavevector \Vector{n} )\Vector{n} 
%   - (\Wavevector \Delta\Vector{n} )\Vector{n} 
%   - (\Wavevector \Vector{n} )\Delta\Vector{n}   
%   + ( \tilde{\xi} - \xi ) \Vector{n} 
%   + \tilde{\xi} \Delta\Vector{n} 
% \\
 \Delta \wavenumber_{2,i} &\approx& 
   \left( \delta_{ij}  - n_i n_j \right) \Delta\wavenumber_j
   +
   \left(
     - \wavenumber_j  n_i 
     - \wavenumber_m n_m \delta_{ij}  
     + \xi \delta_{ij}
   \right) \Delta n_j
   + n_i \Delta\xi 
\end{eqnarray}
The explicit form of $\Delta\xi = \tilde{\xi} - \xi$ can be found once the material dispersion is specified.
One can even more simplify this equation by assuming an explicit $\Vector{r}_{pilot}$, an explicit $\Wavevector$, and an implicit $\Vector{r}_{pilot}$ dependence in terms
of the surface normal vector $\Vector{n}$ of $\xi$. Therefore 
$\Delta \xi = \left(\partial_k \xi + \frac{\partial \xi}{\partial n_j} \partial_k n_j\right) \Delta r_k + \tfrac{\partial \xi}{\partial k_i} \Delta k_i$. 
Further it is possible to express the $\Vector{n}$ derivatives in terms of $\Delta \Vector{r}$ by describing the surface via
a scalar function $F(\Vector{r}_{pilot}) = 0$. This leads by using
\begin{eqnarray}
 n_i &=& \frac{\partial_i F}{\sqrt{(\partial_k F)(\partial_k F)}}\,,
\end{eqnarray}
to
\remark{In terms of effective local curvature radii?}
\begin{eqnarray}
  \partial_j n_i &=& \frac{\partial_j \partial_\ell F}{\sqrt{(\partial_k F)(\partial_k F)}} \left(\delta_{\ell i} - \frac{\partial_i F \partial_\ell F}{(\partial_k F)(\partial_k F)}\right)\,,\\
    &=& \frac{\partial_j \partial_\ell F}{\sqrt{(\partial_k F)(\partial_k F)}} \left(\delta_{\ell i} - n_\ell n_i\right)\,.\label{eq:nderivative}
\end{eqnarray}
Hence by introducing $a$ and $a+1$ as the respective surface indices we may obtain a matrix formulation for the different delta parts
\begin{eqnarray}
 \Delta r_{a+1\,i} &=& \Delta r_{a\,i}\,,\\
 \Delta \wavenumber_{a+1\,i} &=& 
   \left( \delta_{ij}  - n_i n_j + n_i \frac{\partial \xi}{\partial \wavenumber_{a\,j}}\right) \Delta\wavenumber_{a\,j} \nonumber\\
   &+&
   \left[
      \left(
     - \wavenumber_{a\,j}  n_i 
     - \wavenumber_{a\,m} n_m \delta_{ij}  
     + \xi \delta_{ij}
     + n_i \frac{\partial \xi}{\partial n_j}
    \right) \partial_\ell n_j + \partial_\ell \xi\right]\Delta r_{a\,\ell}\,.
\end{eqnarray}


\subsection{Isotropic Media}
We assume a scalar, complex valued permittivity $\scalarpermittivity_2$.
\begin{eqnarray}
 \xi &=& \sqrt{\omega^2 \vacuumpermeability \scalarpermittivity_2 - \Wavevector_{\parallel} \cdot \Wavevector_{\parallel}} \nonumber\\
      &=& \sqrt{\omega^2 \vacuumpermeability \scalarpermittivity_2 - (\Wavevector - \scpm{\Wavevector}{\Vector{n}}\Vector{n}) \cdot (\Wavevector - \scpm{\Wavevector}{\Vector{n}}\Vector{n})}\\
 %\Delta \xi &=& \sqrt{\omega^2 \vacuumpermeability \scalarpermittivity_2 - \tilde{\Wavevector}_{\parallel} \cdot \tilde{\Wavevector}_{\parallel}} - \sqrt{\omega^2 \vacuumpermeability \scalarpermittivity_2 - \Wavevector_{\parallel} \cdot \Wavevector_{\parallel}} \\ 
% \tilde{\Wavevector}_{\parallel} &\approx& \Wavevector_\parallel + \Delta\Wavevector - (\Delta\Wavevector \Vector{n})\Vector{n} - (\Wavevector\Vector{n})\Delta\Vector{n} - (\Wavevector \Delta\Vector{n})\Vector{n} \\
% \tilde{\wavenumber}_{\parallel, i} &\approx& \wavenumber_{\parallel,i} + \Delta\wavenumber_i - (\Delta\wavenumber_j n_j) n_i - (\wavenumber_j n_j) \Delta n_i - (\wavenumber_j \Delta n_j) n_i \\
% \tilde{\wavenumber}_{\parallel, i} \cdot \tilde{\wavenumber}_{\parallel, i} 
%   &\approx& \wavenumber_{\parallel,i} \cdot \wavenumber_{\parallel,i} + 
%     2 \wavenumber_{\parallel,i} ( \Delta\wavenumber_i - (\Delta\wavenumber_j n_j) n_i - (\wavenumber_j n_j) \Delta n_i - (\wavenumber_j \Delta n_j) n_i )
% \\
% \tilde{\wavenumber}_{\parallel, i} \cdot \tilde{\wavenumber}_{\parallel, i} &\approx& \wavenumber_{\parallel,i} \cdot \wavenumber_{\parallel,i} +  2 \mu \\
% \sqrt{x- 2 \mu} &\approx& \sqrt{x} + \left. \frac{\partial \sqrt{x-2\mu}}{\partial \mu} \right|_{0} \mu = \sqrt{x} - \frac{\mu}{\sqrt{x}} \\
% \Delta \xi &\approx& \sqrt{\omega^2 \vacuumpermeability \scalarpermittivity_2 - \Wavevector_{\parallel} \cdot \Wavevector_{\parallel} - 2 \mu } - \sqrt{\omega^2 \vacuumpermeability \scalarpermittivity_2 - \Wavevector_{\parallel} \cdot \Wavevector_{\parallel}} \\ 
% \Delta \xi &\approx& - \frac{\mu}{\xi} \\
 %\Delta \xi &\approx& - \frac{\wavenumber_{\parallel,i}}{\xi} ( \Delta\wavenumber_i - (\Delta\wavenumber_j n_j) n_i - (\wavenumber_j n_j) \Delta n_i - (\wavenumber_j \Delta n_j) n_i )
\end{eqnarray}
The partial derivatives which are necessary to compute the matrix elements are given by\remark{to be checked!}
\begin{eqnarray}
 \partial_k \xi &=& 0\,,\\
 \frac{\partial \xi}{\partial k_i} &=& -\frac{k_i - (k_\ell n_\ell) n_i}{\xi}\,,\\
 \frac{\partial \xi}{\partial n_i} &=& \frac{(k_m n_m) (k_i - (n_m k_m) n_i)}{\xi}\,.
\end{eqnarray}


\subsection{Conic interfaces}

We represent the change of the intersection location $\Delta\Vector{r}$ as function of 2 independent variables $\Delta x$ and $\Delta y$.
Although this difference is not changed by an optical interface.
\begin{eqnarray}
 \Delta\Vector{r} &=&
 \begin{pmatrix}
  \Delta x \\ \Delta y \\ z(x+\Delta x,y+\Delta y) - z(x,y)
 \end{pmatrix}
 \approx 
 \begin{pmatrix}
  \Delta x \\ \Delta y \\ \frac{\partial z}{\partial x} \Delta x + \frac{\partial z}{\partial y} \Delta y
 \end{pmatrix}
\end{eqnarray}
Normal vector (where $x$, $y$ and $z$ live in the vertex coordinate system)
\begin{eqnarray}
 \Vector{n} &=&- \frac{1}{\sqrt{ 1 - c \rho^2 (x^2 + y^2)}} 
  \begin{pmatrix}
   \rho x \\
   \rho y \\
   \rho ( 1 + c ) z(x,y) - 1
  \end{pmatrix}\,.
\end{eqnarray}
To use equation \eqref{eq:nderivative} we have to calculate the Hessian $\partial_i \partial_j F$ of the conic interface.
This can be done by using the implicit form of a conic section (the factor $\tfrac{1}{2}$ was added to remove the factors of two after
differentiation)
\begin{eqnarray}
 F(\Vector{r}) &=& \frac{1}{2}\rho (1+c) z^2 - z + \frac{1}{2} \rho(x^2 + y^2) = 0\,.
\end{eqnarray}
The partial derivatives are given by
\begin{eqnarray}
 \frac{\partial F}{\partial x} &=& \rho x\,,\\
 \frac{\partial F}{\partial y} &=& \rho y\,,\\
 \frac{\partial F}{\partial z} &=& \rho (1+c) z - 1\,,
\end{eqnarray}
and
\begin{eqnarray}
 \frac{\partial^2 F}{\partial x^2} &=& \rho\,, \quad\frac{\partial^2 F}{\partial x \partial y} = 0\,,\\
 \frac{\partial^2 F}{\partial y^2} &=& \rho\,, \quad\frac{\partial^2 F}{\partial y \partial z} = 0\,,\\
 \frac{\partial^2 F}{\partial z^2} &=& \rho (1+c)\,, \quad\frac{\partial^2 F}{\partial z \partial x} = 0\,,
\end{eqnarray}
Therefore the matrix of the normal vector derivatives is given by
\begin{eqnarray}
 (\partial_i n_j) &=& \frac{1}{\sqrt{ 1 - c \rho^2 (x^2 + y^2)}}
\begin{pmatrix}
  \rho & 0 & 0 \\
  0 & \rho & 0 \\
  0 & 0 & \rho(1+c)
\end{pmatrix}\cdot\nonumber\\
&&
\biggl[
\begin{pmatrix}
 1 & 0 & 0 \\
 0 & 1 & 0 \\
 0 & 0 & 1
\end{pmatrix}\nonumber\\
&&\quad
- \frac{1}{1 - c \rho^2 (x^2 + y^2)}\times\nonumber\\
&&\quad\quad
\begin{pmatrix}
 \rho^2 x^2 & \rho^2 x y & \rho x (\rho (1+c) z - 1) \\
 \rho^2 y x & \rho^2 y^2 & \rho y (\rho (1+c) z - 1) \\
 \rho x (\rho (1+c) z - 1) & \rho y (\rho (1+c) z - 1) & 1-\rho^2 (1+c) (x^2+y^2)%\rho^2 (1+c)^2 z^2 - 2 \rho (1+c) z + 1
\end{pmatrix}
\biggr]
\end{eqnarray}


% \begin{eqnarray}
%  \frac{\partial \Vector{n}}{\partial x} &=& \frac{\partial}{\partial x} \left(
% - \frac{1}{\sqrt{ 1 - c (x^2 + y^2)}} 
%   \begin{pmatrix}
%    \rho x \\
%    \rho y \\
%    \rho ( 1 + c ) z(x,y) - 1
%   \end{pmatrix}
%  \right)
% \end{eqnarray}



\end{document}
